fGFPCA dynamic prediction

### 2023-02-07 f-GFPCA.Rmd
update progress report

And this latent funciton can be approximated by fPCA:

@@

where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$. In the fPCA framework, $\boldsymbol{\Gamma}$ is a diagonal matrix and diagonal elements corresponds to eigenvalues.

@@

@@

The entire time grid into 100 small intervals with equal length. The cut is made on the order index of measurement time, but not the actually time itself. Each bin its labeled by its midpoint (5, 15, 25, ..., 995).

@@

<span style="color: red;">When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval. This means that correlation between repeated measures from the same subject is only introduced by the latent function.</span>

@@

@@

- $\hat{\boldsymbol{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
- $\boldsymbol{\Phi}_m$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
- $\hat{\sigma_{\epsilon}}^2$ is the residual variance from FPCA
- $\hat{\boldsymbol{\eta}}_m$ is the estimated latent Gaussian function up to $t_m$
- $\hat{\boldsymbol{f}}_0^m$ is the FPCA mean function up to $t_m$

@@

$$\hat{\boldsymbol{\eta}} = \hat{\boldsymbol{f}}_0+\boldsymbol{\Phi}\hat{\boldsymbol{\xi}}$$

@@

Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. I can think of a few ways to  predict the following track, as below:

### Refit local GLMMs to estimate latent function

1. Local GLMMs: we will be able to estimate random effect of this subject up to mth bin, thus estimate the latent function up to the mth bin ($\hat{\boldsymbol{\eta}}_m$).

2. fPCA: We will have an incomplete function for this new observation. I think fpca.face can handle that.

3. Use the save formulas as in-sample prediction

### Do not estimate latent function

@@

We use the expectation of latent function conditonal on fPCA score to estimate the true latent function, so that the error term can be removed from following calculation:

$$\hat{\eta_s} = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}$$

Here, $\hat{f}_{0}(s)$ is the mean function from fPCA and $\phi(s)$ are eigenfunctions at time s, which is a K $\times$ 1 vector.

Then the data is generated from a binomial distribution:

$$\begin{aligned}
 \boldsymbol{Y}_s & \sim Binomial(p_s) \\
 logit(p_s) & = \hat{\eta}_s = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}
\end{aligned}$$

@@

$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\int_{\boldsymbol{\xi}} \boldsymbol{\xi} p(\boldsymbol{\xi}|\boldsymbol{Y})d\boldsymbol{\xi}$$
$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

@@

$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\frac{\int_{\boldsymbol{\xi}}\boldsymbol{\xi}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

We know that fPCA scores follow a multivariate normal distribution:

@@

Please note that here, we are assuming $\boldsymbol{Y}_s$ is independent from each other given fPCA score. That is, all correlation across time within the same subject is introduced by underlying eigenfunctions.

If we write this down for a single score:

$$E(\xi_k|\boldsymbol{Y})=\frac{\int \xi_k \int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi_{(-k)}}d\xi_k}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

In this case, I ran into the problem of numeric integration being too slow (adaptIntegrate multidimensional). The function value is also very small. Therefore, I turned to MLE instead.

@@

We utilize the same conditional likelihood of scores:

$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

But now we don't need to calculated the marginal distribution of $\boldsymbol{Y}$ anymore (denominator). So we'll get:

$$\begin{aligned}
p(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi}) \\

l(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto logp (\boldsymbol{Y}|\boldsymbol{\xi})+log p(\boldsymbol{\xi})
\end{aligned}$$

@@

It is not easy to get a closed-form solution for the score equation above. We can try to get a numeric solution instead. Below are some visualization of the prediction performances

@@

Posterior mean from Laplace method is the same as posterior mode. In this sense, empirical bayes method and MLE are in fact equivalent.

https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
https://bookdown.org/rdpeng/advstatcomp/laplace-approximation.html

### In-/Out-of-sample prediction of latent function

@@

<span style="color: red;">Doesn't look biased? </span>
<span style="color: red;">Perhaps because the function value range is so small? </span>

@@

With MLE of scores $\hat{\boldsymbol{\xi}}$

@@

- The three components should be mutually independent? 
- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}[\boldsymbol{\Phi\Gamma \Phi^T}+\sigma_{\epsilon}^2\boldsymbol{I}_m]$.
- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?ß
- $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA
- Very different approach from in-sample estimation

Additional questions:

Variance of $\boldsymbol{\xi}$ and random error is in fact given from the FPCA procedure. But we need to variance of mean function for the actual prediction interval.

What happens to the variance of mean function? In practice it is estimated by smoothed population mean at each observation time. It looks like in existing derivations, outcomes are usually assumed to be centered.

@@

- Interval prediction for MLE
- FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?

>>>>

And this latent funciton can be approximated based on Karhunen–Loève expansion:

@@

where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$.

@@

<span style="color: red;"> Question 1: in FPCA framework, $\boldsymbol{\Gamma}$ is a diagonal matrix of eigenvalues. However in my inplemetation, the eigenvalues are much larger than true $\boldsymbol{\Gamma}$. I am not sure why this happened. Does it has to do with the GLMM step, or decreased density when binning observations? </span>

@@

The entire time grid is binned into 100 small intervals with equal length. The cut is made on the order index of measurement time, but not the actually time itself. Each bin its labeled by its midpoint (5, 15, 25, ..., 995).

@@

<span style="color: red;">Question 2: When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval. This means that correlation between repeated measures from the same subject is only introduced by the latent function? </span>

@@

<span style="color: red;">Follow-up on question 1: It looks like the PC function ranges are different from true PC functions. Maybe some of its variation is absorbed by the score, causing $\boldsymbol{\Gamma}$ to be much greater than its true value?</span>

@@

- $\boldsymbol{\hat{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
- $\boldsymbol{\Phi}_m$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
- $\hat{\sigma_{\epsilon}}^2$ is the residual variance from FPCA
- $\hat{\boldsymbol{\eta}}_m$ is the estimated latent Gaussian function up to $t_m$
- $\hat{\boldsymbol{f}}_0^m$ is the FPCA mean function up to $t_m$

@@

$$\boldsymbol{\hat{\eta}} = \hat{\boldsymbol{f}}_0+\boldsymbol{\Phi}\boldsymbol{\hat{\xi}}$$

@@

Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. It seems there are more than one ways to  predict the following track. The first thing coming to mind was in fact to 1) refit local GLMMs to estimate a incomplete latent function; 2) refit local FPCA with this additional incomplete latent function track; and 3) repeat score estimation from in-sample procedure.

However, if we wish to make prediction without re-estimating any model (neither local GLMMs nor FPCA), we need to use the fitted FPCA model (specifically, $\boldsymbol{\hat{f}}_0$, $\boldsymbol{\Phi}$, $\boldsymbol{\hat{\Gamma}}$, $\hat{\sigma}_{\epsilon}$) to estimate the scores for this new observations (out-of-sample scores). I can think of two ways of doing this:

1. Empirical Bayes: similar to in-sample procedure, estimate the posterior mean 
2. MLE, as often used in longitudinal models.

Below I experimented with both procedures.

@@

@@

$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\int \boldsymbol{\xi} p(\boldsymbol{\xi}|\boldsymbol{Y})d\boldsymbol{\xi}$$
$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

@@

$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\frac{\int \boldsymbol{\xi}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}{\int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

We know that FPCA scores follow a multivariate normal distribution:

@@

<span style="color:red;">Follow up on question 2: Please note that here, we are assuming $\boldsymbol{Y}_s$ is independent from each other given fPCA score. That is, all correlation across time within the same subject is introduced by underlying eigenfunctions. </span>

<!-- If we write this down for a single score:  -->

<!-- $$E(\xi_k|\boldsymbol{Y})=\frac{\int \xi_k \int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi_{(-k)}}d\xi_k}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$ -->

It is very difficult to get a closed-form solution of this formula. Therefore, I tried to approximate the integrals with numeric integration (*adaptIntegrate* function from *cubature* package). In this case, I ran into the problem of computation time. Numeric integration was very, very slow, taking roughly an hour for just the denominator of one subject. The function value is also very small.

@@

We maximize the same conditional likelihood of scores (conditional on observations):

$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

But now we don't need to calculated the marginal distribution of $\boldsymbol{Y}$ anymore (denominator), since it would be constant wrt $\boldsymbol{\xi}$:

$$\begin{aligned}
p(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi}) \\
l(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto logp (\boldsymbol{Y}|\boldsymbol{\xi})+log p(\boldsymbol{\xi})
\end{aligned}$$

@@

It is not easy to get a closed-form solution for the score equation above. We can try to get a numeric solution instead (*multiroot* function from *rootSolve* package).

@@

It seems that $E(\boldsymbol{\xi}|\boldsymbol{Y})$ and $p(\boldsymbol{\xi}|\boldsymbol{Y})$ can be connected through Laplace approximation. Specifically, posterior mean estimated from Laplace method is the same as posterior mode. In this sense, the two methods above are essentially equivalent.

Below I have experimented with *LaplaceDemon::LaplaceApproximation*: https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html

<!-- https://bookdown.org/rdpeng/advstatcomp/laplace-approximation.html -->

### In-/Out-of-sample prediction results of latent function

@@

Here I explored the estimated scores, comparing them with the true scores from simulation setup. The figures actually didn't look biased to me. Perhaps it is an artifact of small functional range. Need to think about that.

@@

<span style="color:red;"> Questions 3: a lot of problems on this part </span>

With an estimate of scores $\hat{\boldsymbol{\xi}}$

@@

- The three components should be mutually independent? 
- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}[\boldsymbol{\Phi\Gamma \Phi^T}+\sigma_{\epsilon}^2\boldsymbol{I}_m]$.
- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?
- $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA
- May need to circle back to in-sample estimation and dig into where that formula came from

@@

- Interval prediction for MLE
- Bias of PC scores
- FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
